Proportional multi-state multiple-cohort life table model

### 2018-04-03 MSLT/MSLTdoc.Rmd
Added missing refs

The model was developed by Jan Barendregt and colleagues  and has been widely used in Australia and New Zealand [@RN38].

@@

In this model we use the term ‘health-adjusted life year’ (HALY). As ‘summary measure of population health’ it measures both quantity and quality of life, where one HALY represent the equivalent of one year in full health (which could be two years with a quality of life of 0.5, for example). Specific types of HALY are the quality-adjusted life year (QALY) and the disability-adjusted life year (DALY). The QALY derives from economics and was first used in the 1960s as a measure of health gain (ref Gold MR, Stevenson D, Fryback DG. HALYs and QALYs and DALYs, oh my: similarities and differences in summary measures of population health. Annual review of public health. 2002;23(1):115-34.). The disability-adjusted life-year (DALY) was developed for use in burden of disease studies as a measure of health loss due to disease (Gold et al.). Our calculated HALYs are neither QALYs not DALYs, but something in between. They are similar to QALYs in that they represent health gains. However, the main difference is in the calculation of the health-related quality of life component. QALYs use measures of utility weights that traditionally represent individual experiences of health, whereas our estimated HALYs use disability weights linked to specific diseases, which were developed for the Global Burden of Disease study (GOld et al). As discussed in past research (REF Roux and Cobiac 2009) the main advantage of using disability weights over utility weights is that disability weights refer to specific diseases rather than health states. We opted to use the more general terms HALYs given that the use of the DALYs terminology may lead to think that our calculations are similar to those in burden of diseases studies (REF Murray 1997). In our study, our model does not explicitly separate years of life lost (YLL) and years lived with disability (YLD) components, but instead calculates the total number of life years lived, adjusted for the average health-related quality of life in those years (by age and sex). In burden of disease studies, DALYs are defined as the sum Years of Life Lost (YLL) and Years Lived with Disability (YLD).

@@

The PMSLT similar to ITHIM is a comparative risk assessment approach (REF BRIGGS 2016) that consist of calculating the change in the health burden for a population of interest from a change in exposure to health risks factors (e.g. physical inactivity, air pollution and road trauma).As depicted in Figure 1, both methods need estimates of the potential impact fraction (PIF), which indicates the proportion of the disease burden attributable to a risk factor of interest (e.g. physical inactivity). A step further back, is the development of scenarios that bring about change in the distribution of the risk factor of interest. For now, we only focus on calculations from the PIF onward, and provide a hypothetical example of change in population levels of physical activity. Incorporation of additional  health risk factor (air pollution, road trauma, NO2 and noise) will be discussed in the relevant code sections.

@@

* **Interaction between multiple diseases**
    The *PMSLT* accounts for the interaction between multiple diseases, with               proportions of the population being able to be in more than one health state (REF      BRIGGS 2016). This avoids  overestimation of outcomes as a result of summing health     outcomes attributable to each disease individually as done in *ITHIM*. It is           important to note that the *PMSLT* assumes that diseases are independent of each       other. That is to say, developing a disease is unrelated to a concurrent               diagnoses of another disease).

* **Mortality rate** 
    The *PMSLT* calculations for changes in life years (and health-adjusted life years)     and mortality outcomes is based on observed mortality rates for the population of      interest. In the *ITHIM* model, if burden of disease estimates from the Global B       Burden of Disease (GBD) study are used, then, the mortality component is based on      the highest attained life expectancy observed in the world.

@@

In what follows, first, we specify input parameters. Second, we present the code with explaining notes. Third, we present examples of outcomes and lastly we comment on topics related to implementation. Here we only included the physical activity health pathway. In the comments section, implementation of exposure to air pollution and road trauma is discussed. Note that in the presentation of input parameters, those needed to calculate PIFs are excluded, as these are common to the ITHIM, expect if trends are included (refer to comments section)

>>>>

The model was developed by Jan Barendregt and colleagues and has been widely used in Australia and New Zealand [@RN38; @RN8299].

@@

In this model we use the term ‘health-adjusted life year’ (HALY). As ‘summary measure of population health’ it measures both quantity and quality of life, where one HALY represent the equivalent of one year in full health (which could be two years with a quality of life of 0.5, for example). Specific types of HALY are the quality-adjusted life year (QALY) and the disability-adjusted life year (DALY). The QALY derives from economics and was first used in the 1960s as a measure of health gain [@RN8158]. The disability-adjusted life-year (DALY) was developed for use in burden of disease studies as a measure of health loss due to disease [@RN8158]. Our calculated HALYs are neither QALYs not DALYs, but something in between. They are similar to QALYs in that they represent health gains. However, the main difference is in the calculation of the health-related quality of life component. QALYs use measures of utility weights that traditionally represent individual experiences of health, whereas our estimated HALYs use disability weights linked to specific diseases, which were developed for the Global Burden of Disease study [@RN8158]. As discussed in past research [@RN21; @RN8262] the main advantage of using disability weights over utility weights is that disability weights refer to specific diseases rather than health states. We opted to use the more general terms HALYs given that the use of the DALYs terminology may lead to think that our calculations are similar to those in burden of diseases studies [@RN8153]. In our study, our model does not explicitly separate years of life lost (YLL) and years lived with disability (YLD) components, but instead calculates the total number of life years lived, adjusted for the average health-related quality of life in those years (by age and sex). In burden of disease studies, DALYs are defined as the sum Years of Life Lost (YLL) and Years Lived with Disability (YLD).

@@

The PMSLT similar to ITHIM is a comparative risk assessment approach [@RN8957] that consist of calculating the change in the health burden for a population of interest from a change in exposure to health risks factors (e.g. physical inactivity, air pollution and road trauma).As depicted in Figure 1, both methods need estimates of the potential impact fraction (PIF), which indicates the proportion of the disease burden attributable to a risk factor of interest (e.g. physical inactivity) [@RN42]. A step further back, is the development of scenarios that bring about change in the distribution of the risk factor of interest. For now, we only focus on calculations from the PIF onward, and provide a hypothetical example of change in population levels of physical activity. Incorporation of additional  health risk factor (air pollution, road trauma, NO2 and noise) will be discussed in the relevant code sections.

@@

* **Interaction between multiple diseases**
    The *PMSLT* accounts for the interaction between multiple diseases, with proportions of the population being     able to be in more than one health state [@RN8957]. This avoids  overestimation of outcomes as a result of       summing health outcomes attributable to each disease individually as done in *ITHIM*. It is important to note     that the *PMSLT* assumes that diseases are independent of each other. That is to say, developing a disease is     unrelated to a concurrent diagnoses of another disease).

* **Mortality rate** 
    The *PMSLT* calculations for changes in life years (and health-adjusted life years) and mortality outcomes is     based on observed mortality rates for the population of interest. In the *ITHIM* model, if burden of disease     estimates from the Global Burden of Disease (GBD) study are used, then, the mortality component is based on      the highest attained life expectancy observed in the world.

@@

In what follows, first, we specify input parameters. Second, we present the code with explaining notes. Third, we present examples of outcomes and lastly we comment on topics related to implementation. Here we only included the physical activity health pathway. In the comments section, implementation of exposure to air pollution and road trauma is discussed. Note that in the presentation of input parameters, those needed to calculate PIFs are excluded, as these are common to the ITHIM, expect if trends are included (refer to comments section).
